---
title: "Group 28 presentation"
format: revealjs
author: 
"Iben Sommerand - s2035--"
"Jonas Sennek - s2035--"
"Emilie Wenner - s---"
"Torbjørn Bak Regueira - s2035--"
"Vedis Arntzen - s203546
editor: visual
---

## Introduction

```{r}
#| echo: false
#| eval: true
#| message: false
library(tidyverse)
library(patchwork)
library(scales)
library(ggridges)
library(patchwork)
library(ggseqlogo)
library(dplyr)
library(table1)
library(purrr)
library(broom)
```

## Materials and method

-   The analysis was performed on the dataset "GDC TCGA Breast Cancer (BRCA)"
-   We included data on gene expression RNAseq and phenotype

```{r}
#| output: false
raw_dir <- here("data/_raw/")
data_loc <- "https://gdc-hub.s3.us-east-1.amazonaws.com/download/"
download_file(raw_dir, 
              data_loc, 
              data_file = "TCGA-BRCA.star_fpkm.tsv.gz", 
              new_file_name = "gene_expression_fpkm.tsv.gz")
```

Materials: What data did you use and where did you get it from? Methods: Which modelling did you use? Think of the methods section as a recipe for how to go from raw to results =\> Flow chart?

## Analysis 1: Days to death and .....

## Analysis 2: PCA

## Analysis 3: Linear model

## Analysis 4:

## Discussion
