---
title: "01_data_load"
format: html
editor: visual
---

## Load data

### Downloading data

```{r}
#| output: false
library("here")
library("tidyverse")
source("99_proj_func.R")
set.seed(28)
```

```{r}
#| output: false
raw_dir <- here("data/_raw/")
data_loc <- "https://gdc-hub.s3.us-east-1.amazonaws.com/download/"
download_file(raw_dir, 
              data_loc, 
              data_file = "TCGA-BRCA.star_fpkm.tsv.gz", 
              new_file_name = "gene_expression_fpkm.tsv.gz")
```

```{r}
#| output: false
download_file(raw_dir, 
              data_loc, 
              data_file = "TCGA-BRCA.clinical.tsv.gz", 
              new_file_name = "phenotypes.tsv.gz")
```

```{r}
#| output: false
download_file(raw_dir, 
              data_loc, 
              data_file = "TCGA-BRCA.survival.tsv.gz", 
              new_file_name = "survival.tsv.gz")
```

## Loading data

```{r}
gene_expression <- read_tsv(file = here("data/_raw/gene_expression_fpkm.tsv.gz"), 
                            na = c("NA",""," "))
phenotypes <- read_tsv(file = here("data/_raw/phenotypes.tsv.gz"), 
                       na = c("NA","", " "))
```

```{r}
spec(phenotypes)
```

```{r}
phenotypes |> select(year_of_diagnosis.diagnoses)

```

### Columns to load

### Phenotype

sample, gender.demographic, ajcc_pathologic_stage.diagnoses, age_at_diagnosis.diagnoses, ethnicity.demographic, days_to_death.demographic, vital_status.demographic, treatment_type.treatments.diagnoses, treatment_or_therapy.treatments.diagnoses, primary_diagnosis.diagnoses, prior_malignancy.diagnoses, year_of_diagnosis.diagnoses

To tidy -\> treatment_or_therapy.treatments.diagnoses, treatment_type.treatments.diagnoses

### Gene expression

Take 1000 random genes

Pivot longer

```{r}
phenotypes_subset <- phenotypes |> select(sample, 
                                          gender.demographic, 
                                          ajcc_pathologic_stage.diagnoses, 
                                          age_at_diagnosis.diagnoses, 
                                          ethnicity.demographic, 
                                          days_to_death.demographic, 
                                          vital_status.demographic,
                                          treatment_or_therapy.treatments.diagnoses,
                                          primary_diagnosis.diagnoses, 
                                          prior_malignancy.diagnoses, 
                                          year_of_diagnosis.diagnoses)
```

```{r}
gene_expression_subset <- gene_expression |> sample_n(size = 2000)
```

Pivoting longer

```{r}
gene_expression_subset_long_format <- gene_expression_subset |> pivot_longer(cols = !Ensembl_ID, names_to = "sample", values_to = "gene_expression")

```

Joining the two datasets

```{r}
loaded_data <- gene_expression_subset_long_format |> 
  inner_join(phenotypes_subset, 
             join_by(sample))

loaded_data |> write_tsv(file = here("data/01_dat_load.tsv.gz"))
```
