```{r}
#| output: false
#| echo: false
#| warning: false
library("tidyverse")
library("here")
library("dplyr")
```

## Analysing the data

```{r}
#| output: false
#| echo: false
#| warning: false
data <- read_tsv(file = here("data/03_dat_aug.tsv.gz"), 
                       na = "NA", 
                      guess_max = 1001)
```

## Plot of years at diagnosis vs age at diagnosis, by cancer stage.

The plots highlight the correlation between the age of patient diagnosed and when the patient dies and by which diagnosis (stage).

### Filter and factorize the data
```{r}
test_data <- data |>
  filter(
    !is.na(age_at_diagnosis_years) & 
    !is.na(years_from_diagnosis_to_death)
  ) |>
  mutate(
    cancer_stage = factor(
      cancer_stage,
      levels = c(
        "Stage I", 
        "Stage IA", 
        "Stage IB", 
        "Stage II", 
        "Stage IIA", 
        "Stage IIB", 
        "Stage III", 
        "Stage IIIA", 
        "Stage IIIB",
        "Stage IIIC", 
        "Stage IV"
      )
    )
  )

```
### Calculate correlations for each cancer stage
```{r}
correlations <- test_data |>
  group_by(cancer_stage) |>
  summarize(
    correlation = cor(age_at_diagnosis_years, 
                      years_from_diagnosis_to_death, 
                      use = "complete.obs"),
    .groups = "drop"
  )
```


### Creating the plot

```{r}

test_data |>
  ggplot(aes(x = age_at_diagnosis_years, y = years_from_diagnosis_to_death, color = cancer_stage)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, size = 1.2) + 
  scale_color_manual(
    values = colorRampPalette(c("lightcoral", "darkred"))(9)
  ) +
  facet_wrap(~ cancer_stage, scales = "free") +
  geom_text(
    data = correlations,
    aes(x = Inf, y = Inf, label = paste("cor.r =", round(correlation, 2))),
    inherit.aes = FALSE,
    hjust = 1.2, vjust = 1.5, size = 3
  ) +
  labs(
    title = "Age at Diagnosis vs Years to Death by Cancer Stage",
    x = "Age at Diagnosis (years)",
    y = "Years to Death",
    color = "Cancer Stage"
  ) +
  scale_x_continuous(
    limits = c(20, 90), 
    breaks = seq(20, 90, by = 10)
  ) + 
  scale_y_continuous(
    limits = c(0, 15), 
    breaks = seq(0, 15, by = 2)
  ) + 
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```

```{r}
#| output: false
#| echo: false
#| warning: false
ggsave("../results/years_at_diagnosis_vs_years_to_death_by_stage.png")
```

Separate code for calculation of the correlations and mean age of diagnosis and mean days to death:

```{r}
summary_stats_by_stage <- data |>
  filter(
    !is.na(age_at_diagnosis_years) & 
    !is.na(years_from_diagnosis_to_death)
  ) |> 
  group_by(cancer_stage) |>
  summarize(
    mean_age = mean(age_at_diagnosis_years, na.rm = TRUE),
    mean_years_to_death = mean(years_from_diagnosis_to_death, na.rm = TRUE),
    correlation = cor(age_at_diagnosis_years, 
                      years_from_diagnosis_to_death, 
                      use = "complete.obs"),
    .groups = "drop" 
  )

print(summary_stats_by_stage)

```

### Plot of years at diagnosis vs age at diagnosis.

```{r}
ggplot(data, aes(x = age_at_diagnosis_years, 
                 y = years_from_diagnosis_to_death)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Age at Diagnosis vs Years to Death",
    x = "Age at Diagnosis (years)",
    y = "Years to Death"
  ) +
  theme_minimal()

```
```{r}
#| output: false
#| echo: false
#| warning: false
ggsave("../results/years_at_diagnosis_vs_years_to_death.png")
```


Separate code for calculation of the correlation and mean age of diagnosis and mean years to death:

```{r}
summary_stats <- data |>
  filter(
    !is.na(age_at_diagnosis_years) & 
    !is.na(years_from_diagnosis_to_death)
  ) |>
  summarize(
    mean_age = mean(age_at_diagnosis_years, na.rm = TRUE),
    mean_years_to_death = mean(years_from_diagnosis_to_death, na.rm = TRUE),
    correlation = cor(age_at_diagnosis_years, years_from_diagnosis_to_death, use = "complete.obs")
  )

print(summary_stats)
```

### Setup of linear model to describe years from diagnosis to death

```{r}
filtered_data <- data |>
  filter(
    !is.na(cancer_stage),
    treatment != "not reported"
  )

```
```{r}
lm_model <- lm(
  years_from_diagnosis_to_death ~ 
    cancer_stage + 
    age_at_diagnosis_years + 
    treatment + 
    prior_malignancy,
  data = filtered_data
)

summary(lm_model)
```

### Filter out "not reported" and NA values in `cancer_stage`, and ensure no missing values in key variables
```{r}

filtered_data <- data |>
  filter(
    !is.na(cancer_stage),
    treatment != "not reported",
    !is.na(age_at_diagnosis_years),
    !is.na(years_from_diagnosis_to_death)
  )

filtered_data <- filtered_data |>
  mutate(predicted_years_to_death = predict(lm_model))
```

```{r}

ggplot(filtered_data, aes(x = predicted_years_to_death, 
                          y = years_from_diagnosis_to_death)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, 
              intercept = 0, 
              color = "red", 
              linetype = "dashed") + 
  labs(
    title = "Predicted vs Actual Years to Death",
    x = "Predicted Years to Death",
    y = "Actual Years to Death"
  ) +
  theme_minimal()

```
```{r}
#| output: false
#| echo: false
#| warning: false
ggsave("../results/years_to_death_prediction.png")

```

```{r}
filtered_data |>
  ggplot(aes(x = predicted_years_to_death, 
             y = years_from_diagnosis_to_death, 
             color = treatment)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, 
              intercept = 0, 
              color = "red", 
              linetype = "dashed") +
  labs(
    title = "Predicted vs Actual Years to Death by Treatment",
    x = "Predicted Years to Death",
    y = "Actual Years to Death",
    color = "Treatment"
  ) +
  facet_wrap(~ treatment) + 
  theme_minimal() +
  theme(
    strip.text = element_text(size = 10, 
                              face = "bold"),
    axis.text.x = element_text(angle = 45, 
                               hjust = 1) 
  )

```
```{r}
#| output: false
#| echo: false
#| warning: false
ggsave("../results/actual_vs_predicted_years_to_death.png")

```

### Plot actual vs. predicted, faceted by treatment and prior malignancy
```{r}
filtered_data |>
  ggplot(aes(x = predicted_years_to_death, 
             y = years_from_diagnosis_to_death, 
             color = treatment)) +
  geom_point(alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Predicted vs Actual Years to Death by Treatment and Prior Malignancy",
    x = "Predicted Years to Death",
    y = "Actual Years to Death",
    color = "Treatment"
  ) +
  facet_grid(prior_malignancy ~ treatment) + 
  theme_minimal() +
  theme(
    strip.text = element_text(size = 10, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

```
```{r}
#| output: false
#| echo: false
#| warning: false
ggsave("../results/years_to_death_prediction_by_treatment_and_prior_malignancy.png")

```
