---
title: "08_analysis_4"
format: html
editor: visual
---
```{r}
#| output: false
library("tidyverse")
library("here")
```


## Loading data
```{r}
#| output: false
data <- read_tsv(file = here("data/03_dat_aug.tsv.gz"), 
                       na = "NA", 
                      guess_max = 1001)
```


## Linear model
Let's create a simple linear model using some of the variables from the datset
to see which variables explain the more variance than others.
```{r}
model <- lm(days_to_death ~ 
              cancer_stage + 
              age_at_diagnosis,
            data = data) 
summary(model)
```
It seems all of the variables are significant, but together they don't explain all the variance in the days_to_death column, which also does make sense.

Let's create a more complicated model to see if we can get a higher R-squared value:

```{r}
model <- lm(days_to_death ~ 
              cancer_stage + 
              age_at_diagnosis + 
              prior_malignancy + 
              treatment + 
              primary_diagnosis, 
            data = data) 
summary(model)
```
This model seem to explain more variance than the first model, however it still only explains a little part of the full variance of the days_to_death column
